---
title: '4329: RSession2'
output:
  html_document: 
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Long-run relationships in finance and multivariate time series

- Cointegration
- Vector Error Correction Models

Modeling Volatility

- ARCH, GARCH


# Testing for Cointegration

```{r}
data = read.csv('https://people.orie.cornell.edu/davidr/SDAFE/data/yields2009.csv')
data = data[, c(3, 4, 5, 6, 7)]
data = as.ts(data)

plot(data, nc = 2)

```

## Philips-Oularis test: po.test (package tseries)

- Testing for Cointegration in Regression (Lecture 3, p20)
- The Phillips-Ouliaris test: regress X1 on other variables and perform Phillips-Perron test for the residual.
- $H_0:$ unit root in cointegrating regression’s residuals, thus the series are not cointegrated. 
- From the plot, the series follows each other closely and there might be cointegration.

```{r}
reg <- lm(X3.mo ~ ., data = data)
plot(reg$residuals, type = "l")
acf(reg$residuals)

```

- The plot seems stationary, while the ACF is not informative about the stationarity of the series.

```{r, message=F, warning=F}
library(tseries)
```


```{r}
po.test(data)

```

- Conclusion: the residuals are stationary and so the series are cointegrated.
- Implementation note: The unit root is estimated from a regression of the first variable (column) of x on the remaining variables of x without a constant and a linear trend.

## Fit VECM: ca.jo (package urca)

```{r, message=F, warning=F}
library(urca)
```


```{r}
vecm = ca.jo(data)

summary(vecm)

```

- conclusion: cannot reject null hypothesis of $r=0$ since test statistics $<$ critical values.

# ARCH and GARCH

Simulate volatility data using GARCH (lecture 4, p19):

$r_t = \phi_0 + \phi_1 r_{t-1} + u_t$ 

$u_t = \epsilon_t \sigma_t$

$\sigma_t^2 = \omega + \alpha u_{t-1}^2 + \beta \sigma_{t-1}^2$

$\epsilon \sim N(0,1)$

```{r}
# N = 2000
N = 100000
set.seed(123)
e = rnorm(N)
u = matrix(1, nrow = N)
r = matrix(0, nrow = N)
sigma2 = matrix(1, nrow = N)

for (t in 1:(N-1)) {
    u[t] = e[t] * sqrt(sigma2[t])
    sigma2[t+1] = 0.03 + 0.2*u[t]^2 + 0.8*sigma2[t]
    r[t+1] = 0.01 + 0.1*r[t] + u[t]
}


plot.ts(data.frame(sigma2, r))

Box.test(u^2, lag = 5, type = 'Ljung-Box')

```

## plot the simulated data


```{r}
par(mfrow=c(2,2))
plot(e[(N-200):N], type="l", xlab="t", ylab=expression(epsilon), main = "epsilon")
plot(sqrt(sigma2[(N-200):N]), type="l", xlab="t", ylab=expression(sigma[t]), main = "sigma_t")
plot(u[(N-200):N], type="l", xlab="t", ylab="u", main = "u")
plot(r[(N-200):N], type="l", xlab="t", ylab="r",main = "r")

par(mfrow=c(2,2))
acf(u, main = "u", lag.max = 20)
acf(u^2,  main = "u^2", lag.max = 20)
acf(r, main = "r", lag.max = 20)
acf(r^2, main = "r^2", lag.max = 20)

```


## Fit ARCH(1): ugarchfit (package rugarch)

```{r, message=F, warning=F}
library(rugarch)
```


```{r}
# variance.model = list(model = "sGARCH"), sGARCH is vanilla GARCH and is the default.
spec = ugarchspec(
    mean.model = list(armaOrder = c(0, 0)),
    variance.model = list(garchOrder = c(1, 0))
)

model = ugarchfit(spec, r)
coef(model)

```

## Fit GARCH(1,1): ugarchfit (package rugarch)

```{r}
spec = ugarchspec(
    mean.model = list(armaOrder = c(0, 0)),
    variance.model = list(garchOrder = c(1, 1))
)

model = ugarchfit(spec, r)
coef(model)

```

## Fit ARMA(1,0) + GARCH(1,1): ugarchfit (package rugarch)


```{r}
spec = ugarchspec(
    mean.model = list(armaOrder = c(1, 0)),
    variance.model = list(garchOrder = c(1, 1))
)

model = ugarchfit(spec, r)
coef(model)

# Intercept (Phi_0) of AR1
(1-coef(model)["ar1"])*coef(model)["mu"]

show(model)

```

The estimated model is very similar with our simulation (N = 100000):

$r_t = 0.0129 + 0.0964 r_{t-1} + u_t$ 

$u_t = \epsilon_t \sigma_t$

$\sigma_t^2 = 0.0307 + 0.2021 u_{t-1}^2 + 0.7969 \sigma_{t-1}^2$

$\epsilon \sim N(0,1)$

Note: $\mu$ from rugarch is $\Phi(L) (1-L)^d (y_t - \mu_t) = \Theta(L) \epsilon_t$.


## Find the optimal GARCH(p,q)

Say you have already found the optimal ARIMA part (for example ARIMA(1, 0), for the GARCH part, try out different specifications, with for example a loop.

For example, to try out all combinations of p and q up to 2:

```{r}
pq = expand.grid(p=1:2, q=1:2) # Returns the cartesian product of the two vectors
pq

# get the AIC with inforcriteria() (returns a vector, first element: AIC, second: BIC)
AICs = matrix(0, nrow = nrow(pq), ncol = 3)
colnames(AICs) = c('AIC', 'p', 'q')
for (row in 1:nrow(pq)) {
    p = pq[row, 1]
    q = pq[row, 2]
    spec = ugarchspec(
    mean.model = list(armaOrder = c(1, 0)),# The optimal ARMA order, found with for ex. auto.arima
    variance.model = list(garchOrder = c(p, q))
    )
    
    model = ugarchfit(spec, r)
    AICs[row, ] = c(infocriteria(model)[1], p, q) # infocriteria(model)[1]: The AIC of the fitted model
}   
AICs

min = min(AICs[, 'AIC'])
min

# The p and q
AICs[AICs[, 'AIC'] == min]
```

- So the final model found (by AIC) was: ARMA(1, 0) + GARCH(1, 1)
- note: ugarchfit finds the parameters with a numerical optimization procedure, where some element of chance may enter, like the initalization of parameters (starting values), and machine specific tolerance criteria (for example, .Machine$double.eps), that may vary between for example Mac and Windows computers. For a high dimensional model, the procedure is minimizing a high dimensional loss function, and may reach slightly different local optima depending on which computer you run the same code on.


## plot the residual of optimal model

```{r}
par(mfrow=c(2,2))
plot(residuals(model, standardize=FALSE)[1:300])
plot(residuals(model, standardize=TRUE)[1:300])
plot(residuals(model, standardize=FALSE)[1:300]^2)
plot(residuals(model, standardize=TRUE)[1:300]^2)

```

- Standardized residual = residuals(model, standardize=FALSE) / sigma(model)
- The standardized residual seems free from autocorrelation, like white noise (test with Ljung-Box (this output is printed by ugarch), can look at ACF as well)


## Fit iGARCH(1,1): ugarchfit (package rugarch)

```{r}
spec = ugarchspec(
    mean.model = list(armaOrder = c(0, 0)),
    variance.model = list(garchOrder = c(1, 1), model='iGARCH')
)

imodel = ugarchfit(spec, r)
coef(imodel)

```



  
  
  
  



