---
title: '4329: RSession1'
output:
  html_document:
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Here we will cover material from slide Lecture1-2 and some part of Lecture 3 (multivariate time series modeling).

Univariate time series modeling

- AR processes
- MA processes
- ARMA processes
- Forecasting

Long-run relationships in finance and multivariate time series

- Stationarity and Unit-Root Testing
- Structural breaks
- Seasonality effects
- VAR and VARMA models

# Univariate time series modeling

## Simulate data

**Simulate ARMA(1,1)**. $y_t = \phi_0 + \phi_1 y_{t-1} + u_t + \theta_1 u_{t-1}$. Where $u_t \sim WN(0,\sigma^2_u)$ .

```{r}
set.seed(123) # Will give you the same results as in these examples
N = 100
# N = 100000
epsilon = rnorm(n = N)
y = vector(length = N)
y[1] = 5
for (i in 2:N) {
    y[i] = 2 +  0.6 * y[i-1] + 0.2 * epsilon[i-1] + epsilon[i]
}

```


```{r}
plot.ts(y)

```


```{r}
mean(y)
var(y)

```

## Autocorrelogram (acf, pacf)

```{r}
acf(y)
acf(y, lag.max = 10)

```


```{r}
pacf(y)

```

For ARMA model, 

- the ARMA process can alternatively be written both as an infinite-order AR process or an infinite-order MA process
- The ACF and PACF are not informative in determining the order of an ARMA model

## Ljung-Box test (Box.test)

- $H_0$: no serial correlation in the first K lags. $\rho(1) = \rho(2)=... = \rho(K) = 0$. The series is white noise. 
- $H_a$: one or more of $\rho(1) = \rho(2)=... = \rho(K) = 0$ is nonzero.

```{r}
Box.test(y, lag = 5, type = 'Ljung-Box')

```


## Estimating time series process

**Estimating AR processes: arima**

- the three integer components (p, d, q) are the AR order, the degree of differencing, and the MA order.
- fitdf takes into account estimated parameters (degrees of freedom  = n_lags−n_parameters = n_lags-fitdf)

```{r}
model1 = arima(y, order = c(1, 0, 0))
model1
Box.test(residuals(model1), lag = 10, type = 'Ljung-Box', fitdf = 1)
```

**Estimating MA processes: arima**

- argument: order = (0, 0, "MA lag")
```{r}
model2 = arima(y, order = c(0, 0, 1))
model2
Box.test(residuals(model2), lag = 10, type = 'Ljung-Box', fitdf = 1)

```

**Estimating ARMA processes: arima**

- argument: order = c(AR, MA)

```{r}
model3 = arima(y, order = c(1, 0, 1))
model3
Box.test(residuals(model3), lag = 10, type = 'Ljung-Box', fitdf = 2)

```

**Model selection of ARMA processes: auto.arima (library forecast)**

- The ARMA(p,q) model that minimizes the AIC
- The AIC is the same with arima(y, order = c(1, 0, 0))

```{r, message=F, warning=F}
library(forecast)
model4 = auto.arima(y, max.p = 5, max.q = 5, d = 0, ic = 'aic')
model4
Box.test(residuals(model4), lag = 10, type = 'Ljung-Box', fitdf = 3)

```

Using the BIC

```{r}
auto.arima(y, max.p = 5, max.q = 5, d = 0, ic = 'bic')

```

**(See what happens when N (in the y simulation at the beginning) is increased ...)**

## Forecast: predict

Refit the ARIMA(1,1) model to the first 90 observations

```{r}
model5 = arima(y[1:90], order = c(1, 0, 1))
forecast = predict(model5, n.ahead=10)
forecast
```

In a plot, with actual values, forecasts, and 2 standard deviations error bars

```{r}
plot(y, type='b')
lines(seq(from = 90, by = 1, length = 10), forecast$pred, col='red') # The forecast
lines(seq(from = 90, by = 1, length = 10),
      forecast$pred + 1.96*forecast$se, col='blue')
lines(seq(from = 90, by = 1, length = 10),
      forecast$pred - 1.96*forecast$se, col='blue')
```


# Stationarity and unit root testing

**Stationarity and unit root testing: adf.test (package tseries)**

$H_0$: there is a unit root

```{r, message=F, warning=F}
library(tseries)
adf.test(y)
```

Barely significant

**Random walk**

```{r}
epsilon = rnorm(n = N)
y2 = vector(length = N)
y2[1] = 5
for (i in 2:N) {
    y2[i] = 0.1 +  y2[i-1] + epsilon[i]
}
plot.ts(y2)
adf.test(y2)
```


```{r}
plot.ts(diff(y2))
adf.test(diff(y2))

```


# Structural breaks

**Structural breaks: sctest (package strucchange)**


```{r}
y3 = c(1 + 0.4 * 1:50, 3 + 1.5 * 51:100) + rnorm(n = 100)
plot.ts(y3)
```

$H_0:$ the parameters are stable over time, in this case, the intercept is stable overtime.

point = potential Structural breaks.

```{r, message=F, warning=F}
library(strucchange)
sctest(y3 ~ 1 , type = "Chow", point = 51)

```



# VAR and VARMA

**VAR and VARMA models: (package vars)**

```{r, message=F, warning=F}
library(vars)
data("Canada") # Macro data set for canada
head(Canada)
class(Canada)
plot(Canada, nc = 2)
plot(diff(Canada), nc = 2)

```



**Select optimal VAR(p) according to AIC: VARselect**

p = 1 optimal

```{r}
var1 = VARselect(diff(Canada))
var1
```



**Fit VAR(p): VAR**

```{r}
var2 = VAR(diff(Canada), p = 1)
var2
```

**Impulse response: irf**

- One unit shock to e's impact on the remaining variables.
- how long and to what degree a shock to a given equation has on all of the variables in the system

```{r}
# the default is n.ahead = 10
irf(var2, impulse = "e", response = c("e", "prod", "rw", "U"), boot = FALSE)

```

**Variance decomposition: fevd**

- how much of the s-step ahead forecast error variance for each variable is explained by innovations to each explanatory variable $(s = 1, 2,...)$

```{r}
fevd(var2)$e
fevd(var2)$prod

```

**Using R's ar function (with data in ts format):**

```{r}
# If the data had not already been in a time series format:
# Canada = as.ts(Canada)

# Selecting columns from ts objects, use [, 'column_name']:
head(Canada[, 'e'])

# Same as the VAR fitted above
var_ar = ar(diff(Canada), order.max = 1, method = 'ols')
var_ar

# get the residuals with ...$resid (not with residuals)
# residuals(var_ar)
head(var_ar$resid)

# Omit the first missing row with na.omit (one is missing because the VAR order is one)
head(na.omit(var_ar$resid))

# Make a prediction from a new observation (that we just set equal to 0.8 times the mean)
new = var_ar$x.mean * 0.8
# Convert to a `ts` object
new_ts = ts(t(data.frame(new)), start = c(2001,1), end = c(2001, 1), frequency = 4)
# The `Canada` data set's last obervation was = 2000 Q4
new_ts

predict(var_ar, new_ts, n.ahead = 1, se.fit = FALSE)

# The acf, the plots on the off-diagonal are the cross correlations, the ones on the diagonal are the acf
acf(na.omit(var_ar$resid))

# The absolute values of the eigenvalues of phi (eigen)
phi = var_ar$ar[1, , ]
phi

abs(eigen(phi)$values)

# The covariance matrix of the residuals (cov)
cov(na.omit(var_ar$resid))

```


