---
title: '4329: RSession0, Quick Introduction'
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Packages

```{r, message=F, warning=F}
# Download and install packages 
# install.packages('dplyr')

# load and attach add-on packages
library(dplyr)
# # ?, help operator, access to the documentation pages for R functions, data sets, and other objects
# ?dplyr
# # see (the first 10) function within dplyr package
# ls("package:dplyr")[1:10]
# ?group_by
# ?library

```

# Data type

```{r}
# vector
vec_num = c(1,2,3,4) #numeric
class(vec_num)
vec_num = 1:3 #integer
class(vec_num)
vec_char = c("a","b","c") # character
class(vec_char)

# matrix
matrix_a = matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE) 
class(matrix_a)

# data frame
df_a = data.frame(id = c(111,222,333), age = c(20,30,40))
View(df_a)
class(df_a)

# list
list_a = list(vec_num = vec_num, vec_char, matrix_a, df_a)
View(list_a)
class(list_a)
list_a$vec_num
list_a[[1]] # member reference
list_a[1] # list slice
```

Small note: why subset with list_a[[1]] instead of list_a[1]? It does not matter in this example, but the difference may be important in other settings, see the [R Inferno] page 72, section 8.1.54 list subscripting


# Code Snippets

Used for quickly inserting common snippets of code, for example:

- fun (function)
- mat (matrix)
- if, el, and ei (conditional expressions)
- for (for loop)

```{r}
# code snippets example: typing for, then press tab
# for (variable in vector) {
#   
# }

```


# Load and save

```{r}
# check current working directory
# getwd()

# set working directory using the following function
# setwd(dir) # e.g. C:/Users/...

# Load the Monthly and Quarterly Databases for Macroeconomic Research from the Federal Reserve
# download data from web
data = read.csv('https://files.stlouisfed.org/files/htdocs/fred-databases/monthly/current.csv')

# save data to RData format
save(data, file = "data.RData")

# save data to csv format
write.csv(data, file = "data.csv")

# remove the data
rm(data)

# Sometimes, the data is given in RData format
# note: if the data and the code are not in the same directory, we need to specify the specific directory for the data 
load("data.RData")

```

# Working through an example

**Check and prepare the data**.

```{r}
class(data)
head(data)
tail(data)
colnames(data)[1:10]
nrow(data)

# The first and last row are useless, drop it using subsetting
# also just keep the first 10 columns/variables
data = data[-c(1,712), 1:10]
# Same as
# data <- data = data[-c(1,712), 1:10]
# Note: the two assignment operators (= and <-) are equivalent (except when used in a function call)
head(data)

# Convert the date string column to a date variable
data$sasdate = as.Date(data$sasdate, '%m/%d/%Y')
head(data)

# Plot one of the series
# again, choose a subset ([...]), but with a vector (c(...))
plot(data[c('sasdate', 'RPI')])

```


**Regression and lapply**

```{r}
# Regress RPI on the W875RX1 series( 3rd column)
reg = lm(RPI ~ W875RX1, data)
summary(reg)

# An object of class "lm" is a list 
class(reg)
class(summary(reg))
# check the element of reg
names(reg)
# View(reg)

# extract the coefficients
coef(reg)
reg$coefficients
summary(reg)$coefficients

# extract the first 10 residuals
resid(reg)[1:10]
reg$residuals[1:10]
summary(reg)$residuals[1:10]

# Regress RPI on all of the colums, and look at the R square
variables = colnames(data)[-(1:2)]
variables

# use lapply to apply a function to all these variable names
# lapply applies a function to every element in a list/vector, and returns a list
# using anonymous function
regressions = lapply(variables,
                     function(x) lm(as.formula(paste('RPI ~ ', x)), data))

# The first element is of course the same as above:
regressions[[1]] 
regressions[1] 

# The R squares
rsquares = lapply(regressions, function(x) summary(x)$r.squared)
rsquares

# In a more readable output
cbind(Rsq = rsquares, name = variables)

```

**Regression and for loop**

```{r}
rsquares = matrix(0, nrow = 8)
i = 1

for (name in variables) {
    reg = lm(as.formula(paste('RPI ~ ', name)), data)
    rsquares[i] = summary(reg)$r.squared
    i = i + 1
}

# Yet another way
rsquares_another = list()

for (name in variables) {
    reg = lm(as.formula(paste('RPI ~ ', name)), data)
    rsquares_another[[name]] = summary(reg)$r.squared
}

rsquares

unlist(rsquares_another)

```

Since we converted the date column to a date variable, we can subset with R's "date arithmetic", to pick rows after January 2000:

```{r}
head(data[data$sasdate > '2000-1-1', ])

```


# Other resources

Help sites:

- [Stackoverflow]
- [Stats stack exchange] 

An alternative to typset your homework:

- [RMarkdown and knitr]

Free books:

- [The Art of R Programming]
- [R for Data Science] (More "modern"/add-ons aspects of R)

Misc:

- [Google R Style Guide]



[R Inferno]: https://www.burns-stat.com/pages/Tutor/R_inferno.pdf
[Stackoverflow]: https://stackoverflow.com/questions
[Stats stack exchange]: https://stats.stackexchange.com/
[RMarkdown and knitr]: https://rmarkdown.rstudio.com/articles_intro.html
[The Art of R Programming]: http://heather.cs.ucdavis.edu/~matloff/132/NSPpart.pdf
[R for Data Science]: https://r4ds.had.co.nz/
[Google R Style Guide]: https://google.github.io/styleguide/Rguide.html
