---
title: "Seminar 1"
author: Hendro Sugandi
output:
  bookdown::pdf_document2: 
    number_sections: yes
    toc: yes
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# options(scipen=999)
options(digits = 5)

```


# Plan

This seminar explains some topics that are covered in lecture notes 1, 2 and 3.

- Random variable
- Descriptive statistics
- Linear combinations of portfolio weights
- Sharpe ratio and Mean-Variance
- Linear algebra
- Sample estimates


# Random variable

- Random variable:
  - is a function which assigns some quantities to all possible outcomes
  - For a continuous random variable, there is a continuous probability density which captures the relative likelihood that the value of the random variable would equal the point

Example 1

```{r}
## Example 1
## make data frame Example 1 (df_Ex1) 
## make rA and rB variables for illustration of covariance 
df_Ex1 <- data.frame(
  Prob = c(0.25, 0.5, 0.25),
  rA = c(0.69, 0.03, -0.43),
  rB = c(-0.06, 0.13, -0.08)
)

```
  


# Descriptive statistics

## Overview

- Mean
- Variance, standard deviation
  - Measure dispersion of a random variable from its mean
- Covariance, correlation
  - Measure the co-movement between two random variables

## Mean

Expected return is:

$$\text{E}[r] = \sum_{s=1}^S \pi_s \ r_s$$

Where s is scenario, $p_s$ is probability of each scenario, and $r_s$ is return in each scenario (Lecture 1, page 9).

```{r}
## E[r] from Example 1
meanA <- sum(df_Ex1$Prob*df_Ex1$rA)
meanB <- sum(df_Ex1$Prob*df_Ex1$rB)
meanA
meanB

```



## Variance and standard deviation

$$\text{Var}[r] = \text{E}[(r - \text{E}[r] )^2] = \sum_{s=1}^S \pi_s \ (r_s - \text{E}[r])^2$$
$$\sigma = \text{SD}(r) = \sqrt{\text{Var}[r]}$$
Lecture 1, page 10.

Variance also can be defined as

$$\text{Var}[X] = \text{E}[X^2] - \text{E}[X]^2$$


```{r}
## Example 1
## calculate Var[r]
varA <- sum(df_Ex1$Prob*(df_Ex1$rA - meanA)^2)
varB <- sum(df_Ex1$Prob*(df_Ex1$rB - meanB)^2)
varA
varB
## calculate sd[r]
sqrt(varA)
sqrt(varB)

```


## Covariance and correlation

$$\text{Cov}[X,Y] = \text{E}[XY] - \text{E}[X] \  \text{E}[Y]$$

$$\rho =\text{Cor}[X,Y] = \frac{Cov[X,Y]}{\sigma_X \sigma_Y}$$

```{r}
## Covariance of rA, rB = covarAB
meanAB <- sum(df_Ex1$Prob*df_Ex1$rA*df_Ex1$rB)
meanAB - meanA* meanB
covarAB <- meanAB - meanA* meanB

## correlation
covarAB/(sqrt(varA)*sqrt(varB))


```

# Linear combinations of portfolio weights

Suppose we want to combine the two stocks as a portfolio. $w_A$ and $w_B$ are portfolio weights on stocks A and B, respectively. 

$$r_p = w_A \ r_A + w_B \ r_B$$

Assume $w_A=0.15$ and $w_B=0.85$

Some properties of expectation and variance: 

$$\text{E}[aX + bY] = a \  \text{E}[X] + b \  \text{E}[Y]$$
$$\text{Var}[aX+bY] = a^2 \text{Var}[X] + b^2 \text{Var}[Y]+2ab \text{Cov}[X,Y]$$
See lecture 2, page 33.

```{r}
## weight
wA <- 0.15
wB <- 0.85

## portfolio expected return
mean_port <- wA*meanA + wB*meanB
mean_port
# portfolio variance
var_port <- wA^2*varA + wB^2*varB + 2*wA*wB*covarAB
var_port
# portfolio sd
sd_port <- sqrt(var_port)
sd_port


```

# Sharpe ratio and mean-variance

## Sharpe ratio


$$S_i = \frac{\text{E}[r_i^e]}{\sigma_i}$$

where $\text{E}[r_i^e]$ is expected excess return of an asset. $\text{E}[r_i^e] = \text{E}[r_i] - r_f$. Assume $r_f = 0$, then


```{r}
# Sharpe ratio of A
(meanA-0)/sqrt(varA)

# Sharpe ratio of B
(meanB-0)/sqrt(varB)

# Sharpe ratio of portfolio
(mean_port-0)/sd_port

```

## Mean-variance

```{r}
N_obs <- 1000

## create data frame
df_plot <- data.frame(
  w_1 = seq(from = 0, to = 1, length.out = N_obs)
)
# create w_2
df_plot$w_2 <- 1 - df_plot$w_1
# calculate portfolio mean and sd for different value of weights
df_plot$R_mean <- df_plot$w_1 * meanA + df_plot$w_2*meanB
df_plot$R_var <- df_plot$w_1^2*varA + df_plot$w_2^2*varB + 
  2*df_plot$w_1*df_plot$w_2*covarAB
df_plot$R_sd <- sqrt(df_plot$R_var)

## plot expected return and volatility for different value of weights
plot(x = df_plot$R_sd, y = df_plot$R_mean, type = "l",
     ylim=c(0.02,0.08), xlab = "Volatility", ylab = "Expected return")
points(x = sqrt(varA), y = meanA, col = "red")
points(x = sqrt(varB), y = meanB, col = "blue")
points(x = sqrt(var_port), y = mean_port, col = "green")
text(x = c(sqrt(varA), sqrt(varB), sqrt(var_port)), 
     y = c(meanA, meanB, mean_port)-0.002, 
     labels = c("A", "B", "P"), cex= 0.7)


```



# Linear algebra 

## General 

$$\mu = \begin{bmatrix}
E[r_A] \\
E[r_B]  
\end{bmatrix}$$

$$\Sigma = \begin{bmatrix}
Var[r_A] & Cov[r_A,r_B] \\
Cov[r_A,r_B] & Var[r_B] 
\end{bmatrix} = \begin{bmatrix}
\sigma_{r_A}^2 & \sigma_{r_A,r_B} \\
\sigma_{r_A,r_B} & \sigma_{r_B}^2 
\end{bmatrix}$$


```{r}
mu_mat <- matrix(c(meanA, meanB), nrow = 2)
Sigma_mat <- matrix(c(varA, covarAB, covarAB, varB),
                    byrow = TRUE, nrow = 2)

```

$$\mu^T = \mu' = \begin{bmatrix}
E[r_A] & E[r_B]  \\
\end{bmatrix}$$

```{r}
t(mu_mat)

```


## Application on expected return and volatility {#IdLinAl12}

Let $\mathbf{w}$ be a weight matrix:

$$\mathbf{w} = \begin{bmatrix}
w_A \\
w_B  
\end{bmatrix}$$

The expected return of the portfolio is:

$$
\begin{aligned}
  \mathbf{w}' \ \mu  &= E[r_p]\\
  &= E[w_A \ r_A + w_B \ r_B] \\  
  &= w_A \ E[r_A] + w_B \ E[r_B]\\  
\end{aligned}
$$

The portfolio's variance is:

$$
\begin{aligned}
  \mathbf{w}'\Sigma \mathbf{w}  &= Var[r_p]\\
  &= Var[w_A \ r_A + w_B \ r_B] \\  
  &= w_A^2 \ Var[r_A] + w_B^2 \ Var[r_B] + 2 w_A w_B \sigma_{r_A,r_B}\\  
\end{aligned}
$$

See lecture 3, page 27.

```{r}
w_mat <- c(0.15, 0.85)

## E[rp]
t(w_mat) %*% mu_mat

## Var[rp]
t(w_mat) %*% Sigma_mat %*% w_mat


```


# Sample estimate


- Usually, we observe a sample of realized returns for some period of time, $r_1, ..., r_T$. We can estimate the sample mean and standard deviation. 
- Random variable:
  - For a continuous random variable, there is a continuous probability density which captures the relative likelihood that the value of the random variable would equal the point. For example, we assume that the stock returns follow a normal distribution.

$$f(r) = \frac{1}{\sqrt{2 \pi} \sigma}\ \exp\Big( -\frac{(r- \mu)^2}{2\sigma^2} \Big)$$  


Example 2

```{r}

## Example 2
## generate random numbers that follow Multivariate Normal Distribution
## notice that we use mean and sd as inputs for mvrnorm function
## rA and rB are correlated for illustration
library(MASS) ## for mvrnorm function
# define the covariance matrix for simulated data
Sigma <- matrix(c(7, -3, -3, 4),2,2) / 100
# simulate multi-variate normal distribution
set.seed(1) # for replication
df_Ex2 <- mvrnorm(n = 10000, mu = c(5/100, 3/100), Sigma)
## change data type from matrix to data frame
df_Ex2 <- as.data.frame(df_Ex2)
## assign column names to the data frame
colnames(df_Ex2) <- c("rA", "rB")

## distribution of return
hist(df_Ex2$rA, breaks = 10,
     xlab = "Returns", 
     ylim = c(0,1.5),
     main = "Return distribution of simulated data", probability = TRUE)

# Density line
lines(density(df_Ex2$rA), col = "blue", lwd = 2)

```


Sample mean:

$$\bar{r} = \frac{1}{T} \sum_{t=1}^T r_t$$

```{r}
## sample mean 
mean(df_Ex2$rA)

```

Sample variance and standard deviation

$$\hat{\sigma}^2 = \frac{1}{T-1} \sum_{t=1}^T(r_t - \bar{r})^2$$

```{r}
## variance
var(x = df_Ex2$rA)
## standard deviation
sd(x = df_Ex2$rA)
## equivalently, sd = sqrt(var)
sqrt(var(x = df_Ex2$rA))

## using the function from Lecture 1 code and data: visualize.R
# This function returns the unbiased estimator of volatility 
# Note that this is equivalent to using the function "STEDV.S" in Excel
vol.s <-function(x){
        n<-length(as.vector(x))
        x_bar<-sum(x)/n
        # note that the sum of squared deviations is divided by (n-1)
        stdev<-sqrt(sum((x-x_bar)^2)/(n-1))  
        return(stdev)
}

## similar result with before
vol.s(x = df_Ex2$rA)

## using alternative definition, var[x] = E[x^2]-E[x]^2, 
## notice there are slight differences if we expand the digits
## due to differences in denominator
sqrt(mean(df_Ex2$rA^2) - mean(df_Ex2$rA)^2)
## adjusting for the denominator, we get the same results
sqrt((mean(df_Ex2$rA^2) - mean(df_Ex2$rA)^2)*nrow(df_Ex2)/(nrow(df_Ex2)-1) )

```

Sample covariance

```{r}
cov(df_Ex2$rA, df_Ex2$rB)

## to calculate Covariance matrix
cov(df_Ex2)

```


Now we can perform the same matrix operations as in section \@ref(IdLinAl12) to calculate portfolio's mean and volatility, given the portfolio weights.    

# Illustration of covariance

```{r}
## negative covariance
plot(x = df_Ex2$rA, y = df_Ex2$rB , xlab = "rA", ylab = "rB" )

## positive covariance
Sigma <- matrix(c(7, 3, 3, 4),2,2) / 100
df_Ex2 <- mvrnorm(n = 10000, mu = c(5/100, 3/100), Sigma)
df_Ex2 <- as.data.frame(df_Ex2)
colnames(df_Ex2) <- c("rA", "rB")
plot(x = df_Ex2$rA, y = df_Ex2$rB , xlab = "rA", ylab = "rB" )

## zero covariance
Sigma <- matrix(c(7, 0, 0, 4),2,2) / 100
df_Ex2 <- mvrnorm(n = 10000, mu = c(5/100, 3/100), Sigma)
df_Ex2 <- as.data.frame(df_Ex2)
colnames(df_Ex2) <- c("rA", "rB")
plot(x = df_Ex2$rA, y = df_Ex2$rB , xlab = "rA", ylab = "rB" )


```

