---
title: "PCA on yield curve"
output: 
  html_document: 
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


# Some stylized facts

- volatility clustering
- fat tails
- high cross-correlation during crisis
- selection bias of active management
- anomaly: size, value, asset growth, profitability, momentum, low volatility anomaly
- leverage effect

See Cont (2001)
https://www.tandfonline.com/doi/abs/10.1080/713665670

Cochrane
https://www.johnhcochrane.com/business-35150-advanced-investments

# Some properties

- Assets: S&P500, AAPL, exchange rate IDR
- Mean, standard deviation, skewness, kurtosis of returns
- Autocorrelation
- ...

# Download data

```{r}
library(tidyquant) # to use getYieldCurve
library(lubridate)
library(tidyverse)
library(quantmod)

# ##### DOWNLOAD data, or ...
# ## download: S&P500, AAPL, TSLA, KO, PFE
# df_price <- tq_get(c("EIDO", "^GSPC", "AAPL", "TSLA", "KO", "PFE"),
#                          get = "stock.prices",
#                          from = "1982-01-01" )
# 
# ## save
# save(df_price, file = "df_price.RData" )
# ##### end

##### LOAD SAVED DATA
load(file = "df_price.RData" )
##### end

```

# See the prices and returns

# Tail etc.

# Time series model, best fit, ACF
