---
title: "Introduction to R"
author: Hendro Sugandi
output:
  pdf_document: 
    number_sections: yes
    toc: yes
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Packages

```{r, message=F, warning=F}
# see loaded packages
(.packages())

# # see (the first 10) function within stats package
# ls("package:stats")[1:10]

# # ?, help() operator, access to the documentation pages for R functions, 
# # data sets, and other objects
# ?acf

# Download and install packages 
# install.packages('tidyverse')

# load and attach add-on packages
library(tidyverse)

# ?group_by
# ?library
# or use function: help()

```

# Data type

```{r}
# vector
vec_num = c(1,2,3,4) #numeric
class(vec_num)
vec_num = 1:3 #integer
class(vec_num)
vec_char = c("a","b","c") # character
class(vec_char)

# matrix
matrix_a = matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE) 
class(matrix_a)

# data frame
df_a = data.frame(id1 = c(111,222,333), 
                  id2 = c("a", "b", "c"), 
                  age = c(20,30,40))
View(df_a)
class(df_a)

# select or remove 
df_a[2:3,] # select only the second and third rows
df_a[-c(1),] # exclude the first row
df_a[, 1] # select the first column

# list
list_a = list(vec_num = vec_num, vec_char, matrix_a, df_a)
View(list_a)
class(list_a)
list_a$vec_num
list_a[[1]] # member reference
list_a[1] # list slice

class(list_a[[1]] )
class(list_a[1]) 

# or use function: str(), to check data structure
str(list_a)

```

# Code Snippets

Used for quickly inserting common snippets of code, for example:

- fun (function)
- mat (matrix)
- if, el, and ei (conditional expressions)
- for (for loop)

```{r}
# code snippets example: typing "for", then press tab
# for (variable in vector) {
#   
# }

```


# Load and save

```{r}
# check current working directory
# getwd()

# set working directory using the following function
# setwd(dir) # e.g. "C:/Users/.../"

# read csv
FF5 <- read.csv(file = "F-F_Research_Data_5_Factors_2x3.csv")
# FF5 <- read.csv(file = "C:/Users/.../F-F_Research_Data_5_Factors_2x3.csv")
# OR
# FF5 <- read.csv(file = "C:\\Users\\...\\F-F_Research_Data_5_Factors_2x3.csv")

# save data to RData format
save(FF5, file = "FF5.RData")

# save data to csv format
write.csv(FF5, file = "FF5.csv")

# remove the data
rm(FF5)

# load RData
load("FF5.RData")

```

# Vector and matrix

$$X = \begin{bmatrix}
x_{11} & x_{12} & x_{13}\\
x_{21} & x_{22} & x_{23}
\end{bmatrix}$$

```{r}
# create vector and matrix
vec1 <- matrix(c(10, 20), nrow = 2, ncol = 1)
vec2 <- c(10,20)
mat1 <- matrix(c(10, 20, 30, 40, 50, 60), nrow = 2) # default: byrow = FALSE
mat2 <- matrix(c(10, 20, 30, 40, 50, 60), nrow = 2, byrow = TRUE) 

# check the data dimension
dim(vec1)
dim(vec2)
length(vec2)
dim(mat2)

# transpose
t(vec1)
t(vec2)
class(vec2); class(t(vec2))
t(mat2)

# add
vec1+vec2
mat1+mat2

# multiplication with constant
2*vec1
2*mat2

# matrix multiplication
vec2 %*% t(vec2)
vec1 %*% t(vec1)
t(vec1) %*% mat1
mat1 %*% t(mat2)
t(mat1) %*% mat2

# element wise multiplication
mat1 * mat1

```

$$x^t Y x = \begin{bmatrix}
x_{1} & x_{2} 
\end{bmatrix} 
\begin{bmatrix}
y_{11} & y_{12} \\
y_{21} & y_{22} 
\end{bmatrix}
\begin{bmatrix}
x_{1} \\
x_{2}  
\end{bmatrix}$$

```{r}
# FF5 returns are in percentage
FF5 <- read.csv(file = "F-F_Research_Data_5_Factors_2x3.csv")
x_vec <- matrix(c(0.2, 0.8), nrow = 2) 
sigma_mat <- cov(FF5 %>% select(SMB, HML)) # using dplyr
sigma_mat2 <- cov(FF5[, c(3,4)] )

# variance of portfolio
t(x_vec) %*% sigma_mat %*% x_vec
var(FF5$SMB*0.2+FF5$HML*0.8)


```

Inverse

```{r}
solve(sigma_mat)
round(solve(sigma_mat) %*% sigma_mat, 2) # identity matrix

```


# Example 1: summary statistics, function, and plot

- Load and check data
- Calculate summary statistics
- Create function
- Create plot

```{r}
# read data
FF5 <- read.csv(file = "F-F_Research_Data_5_Factors_2x3.csv")
# check first n row 
head(FF5)
head(FF5, n = 10)
tail(FF5)
colnames(FF5)
dim(FF5); nrow(FF5); ncol(FF5)

# check data structure 
str(FF5)
# summary statistics
summary(FF5)

# change returns from % to actual
FF5 <- FF5[,c(2:7)]/100

# ways of calculating mean of each column
colMeans(FF5)
sapply(FF5, mean)
apply(FF5, MARGIN = 2, FUN = mean)
c(mean(FF5$Mkt.RF), mean(FF5$SMB), mean(FF5$HML), 
  mean(FF5$RMW) , mean(FF5$CMA), mean(FF5$RF))
c(mean(FF5[,1]), mean(FF5[,2]), mean(FF5[,3]), 
  mean(FF5[,4]), mean(FF5[,5]), mean(FF5[,6]))

# using dplyr
FF5 %>%
  summarise_all(.funs = ~ mean(.))

# covariance and correlation
round(cov(FF5), 6)
round(cor(FF5), 2)
cor(FF5$Mkt.RF, FF5$SMB)


```

Make average function

```{r}
f_average <- function(x) {
  N <- length(x)
  average <- sum(x)/N
  return(average)
}

sapply(FF5, f_average)
f_average(x = FF5$Mkt.RF)

```


Using dplyr to calculate yearly average of monthly returns 

```{r}
FF5 <- read.csv(file = "F-F_Research_Data_5_Factors_2x3.csv")

library(lubridate) # for ymd function

FF5 <- FF5 %>%
  # change returns from % to actual
  mutate_at(.vars = vars(Mkt.RF:RF),
            .funs = list( ~ . /100) ) %>%
  # make new variables of date_format and date_year
  mutate(date_format = ymd(paste0(date, "01")),
         date_year = year(date_format)) 

# FF5_sum: FF5 summary
FF5_sum <- FF5 %>%
  # calculate average monthly returns for each year
  group_by(date_year) %>%
  summarise_at(.vars = vars(Mkt.RF:RF),
               .funs = list( ~ mean(.))) %>%
  ungroup()

# plot SMB and HML, average monthly returns for each year  
plot(x = FF5_sum$date_year, y = FF5_sum$SMB, type = "l", 
     col = "red", xlab = "date", ylab = "return") 
lines(x = FF5_sum$date_year, y = FF5_sum$HML, col = "blue") 
legend("bottomleft", legend=c("SMB", "HML"),
       col=c("red", "blue"), lty = 1, cex = 0.7)

# plot SMB and HML, monthly returns
plot(x = FF5$date_format, y = FF5$SMB, type = "l", 
     col = "red", xlab = "date", ylab = "return")
lines(x = FF5$date_format, y = FF5$HML, col = "blue") 
legend("bottomleft", legend=c("SMB", "HML"),
       col=c("red", "blue"), lty = 1, cex = 0.7)


```



# Example 2: regression and for loop

- Simulate data
- Run regressions
- Find the true relationship between Y and X

```{r}
set.seed(1) # for replication

N_obs = 1000

# generate random numbers for X
df_regression <- data.frame(
  #Random number UNIForm
  X1 = runif(n = N_obs, min = -5, max = 10) ,
  #Random number NORMal
  X2 = rnorm(n = N_obs, mean = 3, sd = 5), 
  X3 = rexp(n = N_obs, rate = 1),  
  X4 = rbinom(n = N_obs, size = 10, prob = 0.3),
  X5 = rpois(n = N_obs, lambda = 3),
  noise = rnorm(n = N_obs, mean = 0, sd = 1)
)

# define Y
df_regression$Y <- 3*df_regression$X1 + df_regression$noise

# prepare data frame to store output of loop 
df_out <- data.frame(coef = rep(NA, 5),
                     ts = rep(NA, 5),
                     R2 = rep(NA, 5))

# run univariate regressions
for (i in 1:5) {
  # # run regression at each loop, using X_i
  # reg_temp <- lm(df_regression$Y ~ df_regression[, i])
  
  # alternatively
  f <- as.formula(paste0("Y ~ X",i))
  reg_temp <- lm(f, data = df_regression)
  
  # store the results
  df_out[i,] <- c(reg_temp$coefficients[2],
                  summary(reg_temp)$coefficients[2,3],
                  summary(reg_temp)$r.squared)

}

round(df_out,2)


```


# Further readings

- [R for Data Science], free book
- [RMarkdown], for writing reports and interactive documents
- [Stackoverflow], question and answer website for programming


[R for Data Science]: https://r4ds.had.co.nz/
[RMarkdown]: https://rmarkdown.rstudio.com/lesson-1.html
[Stackoverflow]: https://stackoverflow.com/


